---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/", # nolint
  eval = TRUE
)
```

# EpiNow2: Estimate real-time case counts and time-varying epidemiological parameters

[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://lifecycle.r-lib.org/articles/stages.html#maturing) [![R-CMD-check](https://github.com/epiforecasts/EpiNow2/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/epiforecasts/EpiNow2/actions/workflows/R-CMD-check.yaml) [![codecov](https://codecov.io/gh/epiforecasts/EpiNow2/branch/main/graph/badge.svg?token=FZWwEMdpq6)](https://app.codecov.io/gh/epiforecasts/EpiNow2) [![metacran downloads](http://cranlogs.r-pkg.org/badges/grand-total/EpiNow2?color=ff69b4)](https://cran.r-project.org/package=EpiNow2)

[![MIT license](https://img.shields.io/badge/License-MIT-blue.svg)](https://github.com/epiforecasts/EpiNow2/blob/main/LICENSE.md/)   [![GitHub contributors](https://img.shields.io/github/contributors/epiforecasts/EpiNow2)](https://github.com/epiforecasts/EpiNow2/graphs/contributors)  [![universe](https://epiforecasts.r-universe.dev/badges/EpiNow2)](http://epiforecasts.r-universe.dev/ui/#package:EpiNow2) [![GitHub commits](https://img.shields.io/github/commits-since/epiforecasts/EpiNow2/v1.4.0.svg?color=orange)](https://GitHub.com/epiforecasts/EpiNow2/commit/main/) [![DOI](https://zenodo.org/badge/272995211.svg)](https://zenodo.org/badge/latestdoi/272995211)

## Summary

EpiNow2 estimates the time-varying reproduction number, growth rate, and doubling time using a range of open-source tools ([Abbott et al.](https://doi.org/10.12688/wellcomeopenres.16006.1)), and current best practices ([Gostic et al.](https://doi.org/10.1371/journal.pcbi.1008409)). It aims to help users avoid some of the limitations of naive implementations in a framework that is informed by community feedback and is actively supported.

Forecasting is also supported for the time-varying reproduction number, infections, and reported cases using the same generative process approach as used for estimation.

<details> <summary> More details </summary>

EpiNow2 estimates the time-varying reproduction number on cases by date of infection (using a similar approach to that implemented in [`{EpiEstim}`](https://github.com/mrc-ide/EpiEstim)). Imputed infections are then mapped to observed data (for example cases by date of report) via a series of uncertain delay distributions (in the examples in the package documentation these are an incubation period and a reporting delay) and a reporting model that can include weekly periodicity. 

Uncertainty is propagated from all inputs into the final parameter estimates, helping to mitigate spurious findings. This is handled internally. The time-varying reproduction estimates and the uncertain generation time also give time-varying estimates of the rate of growth.

</details>

<details> <summary> Models provided </summary>

`{EpiNow2}` is designed to be used via a single function call to two functions:

* [`epinow()`](https://epiforecasts.io/EpiNow2/reference/epinow.html): Estimate Rt and cases by date of infection and forecast these infections into the future.

* [`regional_epinow()`](https://epiforecasts.io/EpiNow2/reference/regional_epinow.html): Efficiently run `epinow()` across multiple regions in an efficient manner.

These two functions call [`estimate_infections()`](https://epiforecasts.io/EpiNow2/reference/estimate_infections.html), which works to reconstruct cases by date of infection from reported cases.

[`estimate_infections()`](https://epiforecasts.io/EpiNow2/reference/estimate_infections.html) can be used on its own to infer the underlying infection case curve from reported cases and estimate Rt. Estimating the underlying infection case curve via back-calculation (and then calculating Rt) is substantially less computationally demanding than generating using default settings but may result in less reliable estimates of Rt.

The default model in `estimate_infections()` uses a non-stationary Gaussian process to estimate the time-varying reproduction number and then infer infections. Other options include:

* A stationary Gaussian process (faster to estimate but currently gives reduced performance for real time estimates).
* User specified breakpoints.
* A fixed reproduction number.
* As piecewise constant by combining a fixed reproduction number with breakpoints.
* As a random walk (by combining a fixed reproduction number with regularly spaced breakpoints (i.e weekly)).
* Inferring infections using deconvolution/back-calculation and then calculating the time-varying reproduction number.
* Adjustment for the remaining susceptible population beyond the forecast horizon.

These options generally reduce runtimes at the cost of the granularity of estimates or at the cost of real-time performance.

The documentation for [`estimate_infections`](https://epiforecasts.io/EpiNow2/reference/estimate_infections.html) provides examples of the implementation of the different options available. 

### Other functions

* [`estimate_secondary()`](https://epiforecasts.io/EpiNow2/reference/estimate_secondary.html): Estimate the relationship between primary and secondary observations, for example, deaths (secondary) based on hospital admissions (primary), or bed occupancy (secondary) based on hospital admissions (primary).

* [`estimate_truncation()`](https://epiforecasts.io/EpiNow2/reference/estimate_truncation.html): Estimate a truncation distribution from multiple snapshots of the same data source over time. For more flexibility, check out the [`{epinowcast}`](https://package.epinowcast.org/) package.

For more details on using each function see the [function documentation](https://epiforecasts.io/EpiNow2/reference/index.html).

</details>

## Installation

Install the released version of the package:

```{r, eval = FALSE}
install.packages("EpiNow2")
```

Install the development version of the package with:

```{r, eval = FALSE}
install.packages("EpiNow2", repos = "https://epiforecasts.r-universe.dev")
```

Alternatively, install the development version of the package with [pak](https://pak.r-lib.org/)
as follows (few users should need to do this):

```{r, eval = FALSE}
# check whether {pak} is installed
if (!require("pak")) {
  install.packages("pak")
}
pak::pkg_install("epiforecasts/EpiNow2")
```

If using `pak` fails, try:
```{r, eval = FALSE}
# check whether {remotes} is installed
if (!require("remotes")) {
  install.packages("remotes")
}
remotes::install_github("epiforecasts/EpiNow2")
```

Windows users will need a working installation of Rtools in order to build the package from source. See [here](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started#checking-the-c-toolchain) for a guide to installing Rtools for use with Stan (which is the statistical modelling platform used for the underlying model). For simple deployment/development a prebuilt docker image is also available (see documentation [here](https://github.com/epiforecasts/EpiNow2/wiki/Docker)).

## Resources

<details> <summary> Getting Started </summary>

The [Getting Started vignette](https://epiforecasts.io/EpiNow2/dev/EpiNow2.html)
is your quickest entry point to the package. It provides a quick run through of
the two main functions in the package and how to set up them up. It also
discusses how to summarise and visualise the results after running the models.

</details>

<details> <summary> Package website </summary>

The [package website](https://epiforecasts.io/EpiNow2/dev/index.html) provides
various resources for learning about the package, including the function
reference, details about each model (model definition), workflows for
each model (usage), and case studies or literature of applications of
the package.

</details>

<details> <summary> End-to-end workflows </summary>

The [workflow vignette](https://epiforecasts.io/EpiNow2/dev/articles/estimate_infections_workflow.html) provides guidance on the end-to-end process of estimating reproduction numbers and performing short-term forecasts for a disease spreading in a given setting.

</details>

<details> <summary> Model definitions </summary>

On the website, we provide the mathematical definition of each model.
For example, the model definition vignette for `estimate_infections()` can be found [here](https://epiforecasts.io/EpiNow2/dev/articles/estimate_infections.html).

</details>

<details> <summary> Example implementations </summary>

A simple example of using the package to estimate a national Rt for Covid-19 can be found [here](https://gist.github.com/seabbs/163d0f195892cde685c70473e1f5e867).

</details>

<details> <summary> Reporting templates </summary>

Rmarkdown templates are provided in the package (`templates`) for semi-automated reporting of estimates. If using these templates to report your results please highlight our [limitations](https://doi.org/10.12688/wellcomeopenres.16006.1) as these are key to understanding the results from `{EpiNow2}` .

</details>

## Contributing

We welcome all contributions. If you have identified an issue with the package,
you can file an issue [here](https://github.com/epiforecasts/EpiNow2/issues). We also welcome additions and extensions to the underlying model either in the form of options or improvements. If you wish to contribute in any form, please follow the
[package contributing guide](https://github.com/epiforecasts/EpiNow2/blob/main/.github/CONTRIBUTING.md).
