---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/",
  eval = FALSE
)
```

# EpiNow2: Estimate realtime case counts and time-varying epidemiological parameters 

[![R-CMD-check](https://github.com/epiforecasts/EpiNow2/workflows/R-CMD-check/badge.svg)](https://github.com/epiforecasts/EpiNow2/actions)
[![Codecov test coverage](https://codecov.io/gh/epiforecasts/EpiNow2/branch/master/graph/badge.svg)](https://codecov.io/gh/epiforecasts/EpiNow2?branch=master)
[![DOI](https://zenodo.org/badge/247464257.svg)](https://zenodo.org/badge/latestdoi/247464257)

*This package is under development with breaking changes likely.*

This package estimates the time-varying reproduction number, rate of spread, and doubling time using a range of open-source tools and current best practices. It aims to help users avoid some of the limitations of naive implementations in a framework that is informed by community feedback and is under active development. It assumes that only limited data is available on cases by date of onset and instead uses cases by date of report. These are then imputed to case counts by date of infection using an uncertain reporting delay and incubation period via a Gaussian process based method. Right truncation of cases is dealt with internally by `{EpiNow2}`, as is propogating uncertainty from all inputs into the final parameter estimates (helping to mitigate spurious findings). Time-varying estimates of the reproduction number are estimated using a similar approach to that implemented in the [`{EpiEstim}`](https://github.com/annecori/EpiEstim) package by date of infection with a generation time estimate that includes uncertainty and variation over time controlled using a Gaussian process. Time-varying estimates of the rate of growth are derived from the time-varying reproduction estimates and the uncertain generation time. Optionally, the time-varying reproduction number can be forecast forwards in time using an integration with the [`{EpiSoon}`](https://epiforecasts.io/EpiSoon) package and converted to a case forecast using the renewal equation. Alternatively, the time-varying reproduction number and cases can be forecast using a Gaussian process. See the [methods](https://epiforecasts.io/covid/methods.html) section of our Covid-19 site for a detailed discussion of the approach. 
 

## Installation

Install the stable version of the package using [`{drat}`](https://epiforecasts.io/drat/):

```{r, eval = FALSE}
install.packages("drat")
drat:::add("epiforecasts")
install.packages("EpiNow2")
```

Install the development version of the package with: 

```{r, eval = FALSE}
remotes::install_github("epiforecasts/EpiNow2")
```


For simple deployment/development a prebuilt docker image is also available (see documentation [here](https://github.com/epiforecasts/EpiNow2/wiki/Docker)).

## Quick start

*Under development*

See the documentation for [`epinow`](https://epiforecasts.io/EpiNow2/reference/epinow.html), [`estimate_infections`](https://epiforecasts.io/EpiNow2/reference/estimate_infections.html), and [`forecast_infections`](https://epiforecasts.io/EpiNow2/reference/forecast_infections.html). For more on using each function see the [function documentation](https://epiforecasts.io/EpiNow2/reference/index.html). A working implementation for COVID-19 can be found [here](https://github.com/epiforecasts/covid-global/blob/master/update_nowcasts.R). 

### Reporting templates

Rmarkdown templates are provided in the package (`templates`) for semi-automated reporting of the estimates. These are currently undocumented but an example integration can be seen [here](https://github.com/epiforecasts/covid/blob/master/_posts/national/united-kingdom/united-kingdom.Rmd). If using these templates to report your results please highlight our [limitations](https://epiforecasts.io/covid/) as these are key to understanding our results.

## Contributing

File an issue [here](https://github.com/epiforecasts/EpiNow2/issues) if you have identified an issue with the package. Please note that due to operational constraints priority will be given to users informing government policy or offering methodological insights. We welcome all contributions, in particular those that improve the approach or the robustness of the code base.
