---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/",
  eval = TRUE
)
```

# EpiNow2: Estimate real-time case counts and time-varying epidemiological parameters 

[![R-CMD-check](https://github.com/epiforecasts/EpiNow2/workflows/R-CMD-check/badge.svg)](https://github.com/epiforecasts/EpiNow2/actions)
[![Codecov test coverage](https://codecov.io/gh/epiforecasts/EpiNow2/branch/master/graph/badge.svg)](https://codecov.io/gh/epiforecasts/EpiNow2?branch=master)
[![DOI](https://zenodo.org/badge/272995211.svg)](https://zenodo.org/badge/latestdoi/272995211)

This package estimates the time-varying reproduction number, rate of spread, and doubling time using a range of open-source tools ([Abbott et al.](https://doi.org/10.12688/wellcomeopenres.16006.1)), and current best practices ([Gostic et al.](https://doi.org/10.1101/2020.06.18.20134858)). It aims to help users avoid some of the limitations of naive implementations in a framework that is informed by community feedback and is under active development. It estimates the time-varying reproduction number on cases by date of infection (using a similar approach to that implemented in the [`{EpiEstim}`](https://github.com/annecori/EpiEstim)). Imputed infections are then mapped to observed data (for example cases by date of report) via a series of uncertain delay distributions (in the examples in the package documentation these are an incubation period and a reporting delay) and a reporting model that can include weekly periodicity. The default model uses a non-stationary Gaussian process to estimate the time-varying reproduction number but optionally a stationary Gaussian process may be used (faster to estimate but reduced performance for real time estimates) and arbitrary breakpoints can be defined. A fixed reproduction number is also supported. When combined with user defined breakpoints this gives the option of modelling the time-varying reproduction number as piecewise linear. The documentation for [`estimate_infections`](https://epiforecasts.io/EpiNow2/reference/estimate_infections.html) provides examples of the different options available. Propagating uncertainty from all inputs into the final parameter estimates (helping to mitigate spurious findings) is handled internally. Time-varying estimates of the rate of growth are derived from the time-varying reproduction estimates and the uncertain generation time. Optionally, the time-varying reproduction number can be forecast forwards in time using an integration with the [`{EpiSoon}`](https://epiforecasts.io/EpiSoon) package and converted to a case forecast using the renewal equation. Alternatively, the time-varying reproduction number and cases can be forecast using a Gaussian process. As a standalone tool non-parametric back-calculation is also supported using a novel formulation based on a smoothed mean delay shift of reported cases combined with a Gaussian process to determine the most likely outbreak trajectory.
 
## Installation

Install the stable version of the package:

```{r, eval = FALSE}
install.packages("EpiNow2")
```

Install the stable development version of the package using [`{drat}`](https://epiforecasts.io/drat/):

```{r, eval = FALSE}
install.packages("drat")
drat:::add("epiforecasts")
install.packages("EpiNow2")
```

Install the unstable development version of the package with: 

```{r, eval = FALSE}
remotes::install_github("epiforecasts/EpiNow2")
```


Windows users will need a working installation of Rtools in order to build the package from source. See [here](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started#checking-the-c-toolchain) for a guide to installing Rtools for use with Stan (which is the statistical modeling platform used for the underlying model). For simple deployment/development a prebuilt docker image is also available (see documentation [here](https://github.com/epiforecasts/EpiNow2/wiki/Docker)).

## Quick start

`{EpiNow}` is designed to be used with a single function call or to be used in an ad-hoc fashion via individual function calls. In the following section we give an overview of the simple use case. For more on using each function see the [function documentation](https://epiforecasts.io/EpiNow2/reference/index.html). The core functions are: [`epinow`](https://epiforecasts.io/EpiNow2/reference/epinow.html),
[`regional_epinow`](https://epiforecasts.io/EpiNow2/reference/epinow.html),
[`estimate_infections`](https://epiforecasts.io/EpiNow2/reference/estimate_infections.html), and [`forecast_infections`](https://epiforecasts.io/EpiNow2/reference/forecast_infections.html). [`estimate_infections`](https://epiforecasts.io/EpiNow2/reference/estimate_infections.html) can be use on its own to infer the underlying infection case curve from reported cases with Rt optionally returned (on by default). Estimating the underlying infection case curve alone is substantially less computationally demanding than also estimating Rt.

### Reporting delays, incubation period and generation time

Distributions can either be fitted using package functionality or determined elsewhere and then defined with uncertainty for use in `{EpiNow2}`. When data is supplied a subsampled bootstrapped lognormal will be fit (to account for uncertainty in the observed data without being biased by changes in incidence). An arbitrary number of delay distributions are supported with the most common use case likely to be a incubation period followed by a reporting delay.

```{r}
reporting_delay <- EpiNow2::bootstrapped_dist_fit(rlnorm(100, log(6), 1))
## Set max allowed delay to 30 days to truncate computation
reporting_delay$max <- 30

reporting_delay
```

Here we define the incubation period and generation time based on literature estimates for Covid-19 (see [here](https://github.com/epiforecasts/EpiNow/tree/master/data-raw) for the code that generates these estimates).

```{r}
generation_time <- list(mean = EpiNow2::covid_generation_times[1, ]$mean,
                        mean_sd = EpiNow2::covid_generation_times[1, ]$mean_sd,
                         sd = EpiNow2::covid_generation_times[1, ]$sd,
                         sd_sd = EpiNow2::covid_generation_times[1, ]$sd_sd,
                         max = 30)

incubation_period <- list(mean = EpiNow2::covid_incubation_period[1, ]$mean,
                          mean_sd = EpiNow2::covid_incubation_period[1, ]$mean_sd,
                          sd = EpiNow2::covid_incubation_period[1, ]$sd,
                          sd_sd = EpiNow2::covid_incubation_period[1, ]$sd_sd,
                          max = 30)
```

### [epinow](https://epiforecasts.io/EpiNow2/reference/epinow.html) 

This function represents the core functionality of the package and includes results reporting, plotting and optional saving. It requires a data frame of cases by date of report and the distributions defined above. An additional forecasting module is supported via `EpiSoon` and companion packages (see documentation for an example).

Load example case data from `{EpiNow2}`. 

```{r}
reported_cases <- EpiNow2::example_confirmed[1:50]

head(reported_cases)
```

Estimate cases by date of infection, the time-varying reproduction number, the rate of growth and forecast these estimates into the future by 7 days. Summarise the posterior and return a summary table and plots for reporting purposes. If a `target_folder` is supplied results can be internally saved (with the option to also turn off explicit returning of results). *Note that for real use cases more samples and a longer warm up may be needed*.

```{r, message = FALSE, warning = FALSE}
estimates <- EpiNow2::epinow(reported_cases = reported_cases, 
                             generation_time = generation_time,
                             delays = list(incubation_period, reporting_delay),
                             horizon = 7, samples = 1000, warmup = 200, 
                             cores = 4, chains = 4, verbose = TRUE, 
                             adapt_delta = 0.95)

names(estimates)
```

Both summary measures and posterior samples are returned for all parameters in an easily explored format.

```{r}
estimates$estimates
```

Reported cases are returned separately in order to ease reporting of forecasts and model evaluation.

```{r}
estimates$estimated_reported_cases
```

A summary table is returned for rapidly understanding the results and for reporting purposes.

```{r}
estimates$summary
```

A range of plots are returned (with the single summary plot shown below).

```{r, dpi = 330, fig.width = 12, fig.height = 12, message = FALSE, warning = FALSE}
estimates$plots$summary
```


### [Regional epinow](https://epiforecasts.io/EpiNow2/reference/regional_epinow.html)

This function runs the the `epinow` function across multiple regions in an efficient manner.

Define cases in multiple regions delineated by the region variable.

```{r}
reported_cases <- data.table::rbindlist(list(
   data.table::copy(reported_cases)[, region := "testland"],
   reported_cases[, region := "realland"]))

head(reported_cases)
```

Run the pipeline on each region in turn. The commented code (requires the `{future}` package) can be used to run regions in parallel (when in most scenarios `cores` should be set to 1).

```{r, message = FALSE, warning = FALSE}
## future::plan("multisession")
estimates <- EpiNow2::regional_epinow(reported_cases = reported_cases, 
                                      generation_time = generation_time,
                                      delays = list(incubation_period, reporting_delay),
                                      horizon = 7, samples = 1000, warmup = 200,
                                      cores = 4, chains = 4, adapt_delta = 0.95,
                                      verbose = TRUE)
```

Results from each region are stored in a `regional` list with across region summary measures and plots stored in a `summary` list. All results can be set to be internally saved by setting the `target_folder` and `summary_dir` arguments.

Summary measures that are returned include a table formatted for reporting (along with raw results for further processing).

```{r}
estimates$summary$summarised_results$table
```


A range of plots are again returned (with the single summary plot shown below).

```{r, dpi = 330, fig.width = 12, fig.height = 12, message = FALSE, warning = FALSE}
estimates$summary$summary_plot
```

### Reporting templates

Rmarkdown templates are provided in the package (`templates`) for semi-automated reporting of estimates. These are currently undocumented but an example integration can be seen [here](https://github.com/epiforecasts/covid/blob/master/_posts/national/united-kingdom/united-kingdom.Rmd). If using these templates to report your results please highlight our [limitations](https://epiforecasts.io/covid/) as these are key to understanding our results.

## Contributing

File an issue [here](https://github.com/epiforecasts/EpiNow2/issues) if you have identified an issue with the package. Please note that due to operational constraints priority will be given to users informing government policy or offering methodological insights. We welcome all contributions, in particular those that improve the approach or the robustness of the code base.
