---
title: "Understanding delay distributions in EpiNow2"
author: "EpiNow2 Team"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
vignette: >
  %\VignetteIndexEntry{Understanding delay distributions in EpiNow2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

# What delay distributions represent

Infectious disease surveillance data are shaped by delays between events that cannot be directly observed.
The *generation time* is the interval between infection of a primary case and infection of a secondary case it causes.
The *incubation period* is the time from infection to symptom onset.
The *reporting delay* is the time from symptom onset (or another observable event) to the case appearing in surveillance data.

EpiNow2 uses these delay distributions when estimating the time-varying reproduction number and when producing forecasts and nowcasts.
Each delay shifts and smooths the relationship between underlying infections and the observations available to analysts, so getting them right matters for inference.

# Specifying delays

Delays are specified as `dist_spec` objects using the distribution constructors `LogNormal()`, `Gamma()`, `Normal()`, `Fixed()`, and `NonParametric()`.
Parameters can be given as fixed values or with uncertainty expressed through `Normal()` priors on the natural parameters.

```{r distributions, eval=FALSE}
library(EpiNow2)

# Fixed lognormal delay
incubation_period <- LogNormal(
  meanlog = 1.6, sdlog = 0.5, max = 14
)

# Gamma delay with uncertain parameters
reporting_delay <- Gamma(
  shape = Normal(3, 0.5),
  rate = Normal(1, 0.25),
  max = 14
)

# Fixed (delta) distribution for a known constant delay
fixed_delay <- Fixed(value = 3)

# Nonparametric distribution given directly as a PMF
# (zero-indexed: first element is P(delay = 0))
known_pmf <- NonParametric(c(0.0, 0.1, 0.4, 0.3, 0.2))
```

The `max` argument truncates the distribution at a given value, discarding probability mass beyond that point and renormalising.
Alternatively, `cdf_cutoff` can be used to set the truncation point based on a tail probability threshold.

When parameters carry uncertainty, EpiNow2 samples from the prior during fitting, propagating that uncertainty into the final estimates.

# Why naive discretisation is biased

Delay distributions are defined in continuous time, but surveillance data record events in discrete time windows (typically days).
A common approach to discretisation is CDF differencing, where the probability mass for day $d$ is computed as $F(d+1) - F(d)$.
This treats the primary event as though it occurred at a known, exact time, typically the start of the interval.

In practice, both the primary event (e.g. infection) and the secondary event (e.g. symptom onset) are interval-censored.
The primary event could have occurred at any point within its observation window, and the same is true of the secondary event.
CDF differencing ignores this *double censoring* and introduces systematic bias, particularly for short delays where the censoring window is large relative to the delay itself [@placeholder; @placeholder].

# How primarycensored corrects this

EpiNow2 computes delay PMFs using the `primarycensored` package [@placeholder], which accounts for double censoring explicitly.
For each discrete delay $d$, the PMF value is obtained by integrating over all possible primary event times within the primary window and all possible secondary event times within the secondary window.

Concretely, `primarycensored::dprimarycensored()` is called with both a primary window (`pwindow`) and a secondary window (`swindow`), each typically set to 1 day.
For common distribution and primary event combinations (e.g. lognormal delay with uniform primary events), analytical solutions are available, making the computation fast.
When no closed-form solution exists, numerical integration is used as a fallback.

The resulting PMF correctly represents the probability of observing a delay of $d$ days given that both events are recorded in daily intervals.
This is the PMF used internally by EpiNow2's Stan model when computing the likelihood.

# Composing multiple delays

When the observation process involves more than one delay, the individual delay PMFs are convolved to produce a combined delay.
In EpiNow2 this is done by adding `dist_spec` objects together with the `+` operator.

```{r convolution, eval=FALSE}
# Combined delay from infection to report
combined_delay <- incubation_period + reporting_delay
```

The `+` operator signals to the model that these delays should be convolved.
The `c()` function, by contrast, collects independent delay distributions without convolving them (e.g. for separate delay and truncation distributions passed to different parts of the model).

Convolution of nonparametric PMFs is performed directly in discrete space.
For parametric distributions with uncertain parameters, the convolution is handled within the Stan model at each iteration of the sampler.

# Left truncation

Some delays have a natural lower bound that excludes zero.
For example, a generation time of zero days is epidemiologically implausible for most pathogens because transmission requires some minimum period of viral replication.

EpiNow2 supports this through the `max` argument on `bound_dist()` and through the structure of the PMF itself.
A `NonParametric()` distribution can encode left truncation directly by setting early entries to zero.
For parametric distributions, the PMF computation via `primarycensored` respects the support of the underlying continuous distribution, and users can further adjust the PMF to exclude implausible values.

# Fitting delay distributions from data

When line-list data with individual-level event dates are available, `estimate_dist()` can be used to fit a delay distribution that properly accounts for interval censoring and right truncation.
This function uses a Stan model built on `primarycensored` likelihood functions.

```{r estimate-dist, eval=FALSE}
# Fit a lognormal delay distribution from line-list data
# Data should have columns: pdate_lwr, sdate_lwr
# (and optionally pdate_upr, sdate_upr, obs_date)
fit <- estimate_dist(
  data = linelist_data,
  dist = "lognormal"
)
```

The fitted distribution can then be used directly in the modelling pipeline.
For more details see `vignette("estimate-dist")`.

# A practical example

The following brings together the pieces described above into a typical EpiNow2 workflow.

```{r practical, eval=FALSE}
library(EpiNow2)

# Generation time (lognormal, uncertain parameters)
generation_time <- LogNormal(
  meanlog = Normal(1.4, 0.15),
  sdlog = Normal(0.5, 0.05),
  max = 14
)

# Incubation period (fixed parameters)
incubation_period <- LogNormal(
  meanlog = 1.6, sdlog = 0.4, max = 14
)

# Reporting delay (gamma, fixed parameters)
reporting_delay <- Gamma(
  mean = 2, sd = 1, max = 10
)

# Combined delay from infection to observation
delays <- incubation_period + reporting_delay

# Use in estimation
out <- epinow(
  data = example_confirmed,
  generation_time = generation_time_opts(generation_time),
  delays = delay_opts(delays)
)
```

# References
