---
title: "Understanding delay distributions in EpiNow2"
author: "EpiNow2 Team"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
vignette: >
  %\VignetteIndexEntry{Understanding delay distributions in EpiNow2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4,
  fig.path = "delays-"
)
```

# What delay distributions represent

Infectious disease surveillance data are shaped by delays between events that cannot be directly observed.
The *generation time* is the interval between infection of a primary case and infection of a secondary case it causes.
The *incubation period* is the time from infection to symptom onset.
The *reporting delay* is the time from symptom onset (or another observable event) to the case appearing in surveillance data.

EpiNow2 uses these delay distributions when estimating the time-varying reproduction number and when producing forecasts and nowcasts.
Each delay shifts and smooths the relationship between underlying infections and the observations available to analysts, so getting them right matters for inference.

# Specifying delays

Delays are specified as `dist_spec` objects using the distribution constructors `LogNormal()`, `Gamma()`, `Normal()`, `Fixed()`, and `NonParametric()`.
Parameters can be given as fixed values or with uncertainty expressed through `Normal()` priors on the natural parameters.

```{r distributions}
library(EpiNow2)

# Fixed lognormal delay
incubation_period <- LogNormal(
  meanlog = 1.6, sdlog = 0.5, max = 14
)

# Gamma delay with uncertain parameters
reporting_delay <- Gamma(
  shape = Normal(3, 0.5),
  rate = Normal(1, 0.25),
  max = 14
)

# Fixed (delta) distribution for a known constant delay
fixed_delay <- Fixed(value = 3)

# Nonparametric distribution given directly as a PMF
# (zero-indexed: first element is P(delay = 0))
known_pmf <- NonParametric(c(0.0, 0.1, 0.4, 0.3, 0.2))
```

We can visualise any `dist_spec` object using `plot()`.

```{r plot-distributions}
plot(incubation_period)
plot(reporting_delay)
```

The `max` argument right-truncates the distribution at a given value, discarding probability mass beyond that point and renormalising.
Alternatively, `cdf_cutoff` can be used to set the truncation point based on a tail probability threshold.

When parameters carry uncertainty, EpiNow2 samples from the prior during fitting, propagating that uncertainty into the final estimates.

# Why naive discretisation is biased

Delay distributions are often defined in continuous time, but surveillance data record events in discrete time windows (typically days).
EpiNow2 operates at daily resolution, so both observation windows are set to 1 day.
A common approach to discretisation is CDF differencing, where the probability mass for day $d$ is computed as $F(d+1) - F(d)$.
This treats the primary event as though it occurred at a known, exact time, typically the start of the interval.

In practice, both the primary event (e.g. infection) and the secondary event (e.g. symptom onset) are interval-censored.
The primary event could have occurred at any point within its observation window.
CDF differencing ignores this *primary event censoring* and introduces systematic bias, particularly for short delays where the censoring window is large relative to the delay itself [@Charniga2024; @Park2024].

Right truncation also introduces bias: recent events with long delays are systematically missing from surveillance data because the secondary event has not yet been observed.

# How primarycensored corrects this

EpiNow2 computes delay PMFs using the `primarycensored` package [@Park2024], which accounts for primary event censoring explicitly.
The approach integrates the delay CDF over all possible primary event times within the primary window, producing an adjusted CDF.
The PMF is then obtained by differencing this adjusted CDF at successive days.

For common distribution and primary event combinations (e.g. lognormal delay with uniform primary events), analytical solutions are available, making the computation fast.
When no closed-form solution exists, numerical integration is used as a fallback.

The resulting PMF correctly represents the probability of observing a delay of $d$ days given that both events are recorded in daily intervals.
This is the PMF used internally by EpiNow2's Stan model when computing the likelihood.

# Composing multiple delays

When the observation process involves more than one delay, the individual delay PMFs are convolved to produce a combined delay.
In EpiNow2 this is done by adding `dist_spec` objects together with the `+` operator.

```{r convolution}
# Combined delay from infection to report
combined_delay <- incubation_period + reporting_delay
plot(combined_delay)
```

The `+` operator signals to the model that these delays should be convolved.
The `c()` function, by contrast, collects independent delay distributions without convolving them (e.g. for separate delay and truncation distributions passed to different parts of the model).

Convolution of nonparametric PMFs is performed directly in discrete space.
For parametric distributions with uncertain parameters, the convolution is handled within the Stan model at each iteration of the sampler.

# Truncation

Delay distributions must be truncated to a finite support for use in the discrete-time model.
The `max` argument right-truncates the distribution at a given value, discarding probability mass beyond that point and renormalising.

Left truncation excludes delays below a threshold, which is useful when zero-day delays are epidemiologically implausible (e.g. generation times).
A `NonParametric()` distribution can encode this directly by setting early entries to zero.
For parametric distributions, `primarycensored` handles left truncation analytically when computing the adjusted CDF, so the resulting PMF already accounts for the truncation without ad hoc zeroing and renormalising.

# Fitting delay distributions from data

When line-list data with individual-level event dates are available, `estimate_dist()` can be used to fit a delay distribution that properly accounts for interval censoring and right truncation.
For details see `vignette("estimate-dist")`.

# References
