---
title: "Fitting delay distributions with estimate_dist()"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
vignette: >
  %\VignetteIndexEntry{Fitting delay distributions with estimate_dist()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

`estimate_dist()` fits delay distributions from linelist data while accounting for interval censoring and right truncation.
These biases arise routinely in epidemiological data where event times are recorded at daily resolution and recent observations are missing due to reporting delays.
The function uses a Stan model that vendors likelihood functions from the [primarycensored](https://primarycensored.epinowcast.org/) package.
If you use `estimate_dist()`, please cite `primarycensored` in addition to `EpiNow2`.

The fitted distribution is returned as a `<dist_spec>` object, ready for use in other _EpiNow2_ functions such as `estimate_infections()`.

For more flexible delay distribution modelling, see the [epidist](https://epidist.epinowcast.org/) package.

# Set up

We load `EpiNow2` and `primarycensored` (used here only to simulate data).

```{r setup}
library(EpiNow2)
library(primarycensored)
options(mc.cores = 4)
```

# Simulating censored delay data

We simulate a linelist of 200 observations from a lognormal delay distribution with known parameters, using `rprimarycensored()` to generate delays that are subject to daily censoring and right truncation.

```{r simulate}
set.seed(42)
n <- 200
D <- 30

# True parameters
true_meanlog <- 1.5
true_sdlog <- 0.75

# Generate primary event dates over a two-week window
pdate_lwr <- as.Date("2023-01-01") + sample(0:13, n, replace = TRUE)

# Simulate censored, truncated delays
delays <- rprimarycensored(
  n = n,
  rdist = rlnorm,
  meanlog = true_meanlog,
  sdlog = true_sdlog,
  pwindow = 1,
  D = D
)

# Build a date-based linelist
# obs_date is set per-observation so that each primary event
# has the correct truncation window
linelist <- data.frame(
  pdate_lwr = pdate_lwr,
  sdate_lwr = pdate_lwr + delays,
  obs_date = pdate_lwr + D
)

head(linelist)
```

The data frame has three columns:

- `pdate_lwr`: lower bound of the primary event date (e.g. date of infection).
- `sdate_lwr`: lower bound of the secondary event date (e.g. date of symptom onset).
- `obs_date`: the date at which the data were extracted, used to determine the right truncation point for each observation.

Upper bounds for the primary and secondary event windows default to one day after the lower bounds (daily censoring).

# Fitting the model

We pass the linelist directly to `estimate_dist()`.
The function converts dates to numeric intervals, aggregates identical delay–censoring combinations, and fits the model via MCMC.

```{r fit}
result <- estimate_dist(
  linelist,
  dist = "lognormal",
  verbose = TRUE
)
result
```

The result is a `<dist_spec>` object summarising the posterior mean and standard deviation for each parameter.

# Checking parameter recovery

We can compare the fitted parameters to the true values used in the simulation.

```{r recovery}
fitted_meanlog <- result$parameters$meanlog$parameters$mean
fitted_sdlog <- result$parameters$sdlog$parameters$mean

comparison <- data.frame(
  parameter = c("meanlog", "sdlog"),
  true = c(true_meanlog, true_sdlog),
  fitted = c(fitted_meanlog, fitted_sdlog)
)
comparison
```

With 200 observations the fitted values should be close to the true parameters, though some posterior uncertainty is expected.

# Convenience path for numeric vectors

If you already have a vector of integer-valued delays (e.g. rounded observed delays) rather than a date-based linelist, you can pass it directly.
The function assumes daily censoring and infers a conservative truncation time from the data.

```{r numeric-path}
result_numeric <- estimate_dist(
  delays,
  dist = "lognormal"
)
result_numeric
```

The date-based interface is preferred when primary event dates and observation dates are available, as it allows per-observation truncation times and non-unit censoring windows.
