---
title: "Model overview"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
vignette: >
  %\VignetteIndexEntry{Model overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

EpiNow2 provides several estimation models that can be combined
for end-to-end epidemiological inference and forecasting.
This vignette gives an overview of how these models connect.
For a feature-by-feature reference see the
[model features](model_features.html) vignette.

# Architecture

The diagram below shows the main estimation functions and how
they relate to one another.

```{r architecture, echo = FALSE, fig.alt = "Diagram showing how the main EpiNow2 models connect."}
knitr::include_graphics(
  "model-overview-architecture-1.png"
)
```

Data flows from top to bottom.
`estimate_truncation()` feeds into `estimate_infections()`,
which in turn feeds into `estimate_secondary()` and the
forecasting functions.

# Estimation models

## `estimate_infections()`

The core model estimates latent infections and the
time-varying reproduction number from a count time series.
It supports mechanistic (renewal equation) and
non-mechanistic (backcalculation) approaches.
Configuration is handled by `rt_opts()`, `gp_opts()`,
`gt_opts()`, `delay_opts()`, `obs_opts()`, `trunc_opts()`,
and `backcalc_opts()`.
See the [model definition](estimate_infections.html) and
[configuration examples](estimate_infections_options.html).

## `estimate_secondary()`

Estimates secondary outcomes (e.g. deaths, hospitalisations)
from primary case estimates.
Supports both incidence and prevalence models, configured
via `secondary_opts()`.
See the [model definition](estimate_secondary.html).

## `estimate_truncation()`

A nowcasting model that estimates reporting truncation
distributions from multiple snapshots of the same data at
different reporting dates.
The model itself produces a nowcast (predicted complete
observations), accessible via `get_predictions()`.
Passing the estimated truncation distribution to
`estimate_infections()` via `trunc_opts()` enables
truncation-adjusted inference in the full model.
See the [model definition](estimate_truncation.html).

## `estimate_dist()`

Fits delay distributions from linelist data, accounting for
double interval censoring and right truncation.
Uses likelihood functions vendored from the
[primarycensored](https://primarycensored.epinowcast.org/)
package.
See [fitting delay distributions](estimate-dist.html).

# Simulation

`simulate_infections()` generates observations from a given
Rt trajectory via the renewal equation with fixed parameters.
`simulate_secondary()` does the same for secondary
observations from primary data.
These are useful for model checking and scenario analysis.

# How models chain together

`estimate_truncation()` produces both a nowcast and a
truncation distribution.
The distribution is typically passed to
`estimate_infections()` via `trunc_opts()` for
truncation-adjusted inference, but can also be used via
`delay_opts()` or `gt_opts()` where appropriate.

`estimate_dist()` fits delay distributions from linelist
data.
Its output can define priors for the other models via
`delay_opts()`, `gt_opts()`, or `trunc_opts()`.

The posterior from `estimate_infections()` feeds into
`forecast_infections()`, which projects forward using the
estimated reproduction number trajectory.
`simulate_infections()` can use posterior Rt for scenario
analysis.

Estimated primary observations from `estimate_infections()`
are used as input to `estimate_secondary()` and
`forecast_secondary()`.

`forecast_secondary()` generates forecasts from a fitted
secondary model.
`simulate_secondary()` generates synthetic secondary
observations from primary data.

# Wrapper functions

`epinow()` wraps `estimate_infections()` with logging,
output saving, and optional forecasting.
`regional_epinow()` runs `epinow()` across multiple regions
in parallel.
See [production use](epinow.html).

# Options functions

The table below summarises the options functions, what they
configure, and which estimation functions use them.

| Function | Configures | Default | Used by |
|----------|-----------|---------|---------|
| `rt_opts()` | Reproduction number model | GP on log Rt differences | `estimate_infections()` |
| `gp_opts()` | Gaussian process prior | Matern 3/2 kernel | `estimate_infections()` |
| `gt_opts()` | Generation time distribution | Fixed(1) (no generation time) | `estimate_infections()` |
| `delay_opts()` | Reporting delay distributions | Fixed(0) (no delay) | `estimate_infections()`, `estimate_secondary()` |
| `obs_opts()` | Observation model | Negative binomial with day-of-week effect | `estimate_infections()`, `estimate_secondary()` |
| `trunc_opts()` | Truncation distribution | Fixed(0) (no truncation) | `estimate_infections()` |
| `backcalc_opts()` | Backcalculation settings | Smooth via GP | `estimate_infections()` |
| `secondary_opts()` | Secondary model type | Incidence | `estimate_secondary()` |

See the [prior choice guide](prior_choice_guide.html) for
default values and guidance.

# Documentation map

| Topic | Vignette | Description |
|-------|----------|-------------|
| Getting started | [EpiNow2](EpiNow2.html) | Quick introduction and basic usage |
| Model features | [Model features](model_features.html) | Feature reference with summary and detail tables |
| Infection model | [estimate_infections()](estimate_infections.html) | Mathematical model definition |
| Secondary model | [estimate_secondary()](estimate_secondary.html) | Mathematical model definition |
| Truncation model | [estimate_truncation()](estimate_truncation.html) | Mathematical model definition |
| Delay fitting | [estimate_dist()](estimate-dist.html) | Fitting delay distributions from linelist data |
| Gaussian process | [GP details](gaussian_process_implementation_details.html) | Kernel functions and spectral approximation |
| Practical workflow | [Workflow](estimate_infections_workflow.html) | End-to-end estimation and forecasting |
| Configuration examples | [Options](estimate_infections_options.html) | Different model configurations with results |
| Prior guidance | [Prior choice guide](prior_choice_guide.html) | Default priors and how to modify them |
| Production use | [epinow()](epinow.html) | Using `epinow()` and `regional_epinow()` |
| Case studies | [Case studies](case-studies.html) | External applications in the literature |
