---
title: "Model overview"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
vignette: >
  %\VignetteIndexEntry{Model overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

EpiNow2 provides several estimation models that can be combined for end-to-end epidemiological inference and forecasting.
This vignette gives an overview of how these models connect.
For a reference of what each model can do see the [model features](model_features.html) vignette.

# Architecture

The diagram below shows the main functions and how they relate to one another.

```{r architecture, echo = FALSE, fig.alt = "Diagram showing how the main EpiNow2 models connect."}
knitr::include_graphics(
  "model-overview-architecture-1.png"
)
```

Data flows from top to bottom.
Solid arrows show direct dependencies; dashed arrows show optional connections.
Green boxes are options functions that configure the estimation models.
See the [model features](model_features.html) vignette for what each function and option does.

# How models chain together

`estimate_dist()` fits delay distributions from linelist data, accounting for double interval censoring and right truncation.
Its output can define priors for the other models via `delay_opts()`, `gt_opts()`, or `trunc_opts()`.

`estimate_truncation()` produces both a nowcast and a truncation distribution from multiple snapshots of the same data.
The distribution is typically passed to `estimate_infections()` via `trunc_opts()` for truncation-adjusted inference, but can also be used via `delay_opts()` or `gt_opts()` where appropriate.

`estimate_infections()` is the core model, estimating latent infections and the time-varying reproduction number from a count time series.
Its posterior feeds into `forecast_infections()` for projections and can inform `simulate_infections()` for scenario analysis.

Estimated primary observations from `estimate_infections()` are used as input to `estimate_secondary()`, which estimates secondary outcomes (e.g. deaths, hospitalisations).
`forecast_secondary()` extends a fitted secondary model with new primary data; `simulate_secondary()` generates synthetic secondary observations.

`epinow()` wraps `estimate_infections()` with logging and formatted output.
`regional_epinow()` runs `epinow()` across regions in parallel.

# Documentation map

| Topic | Vignette | Description |
|-------|----------|-------------|
| Model features | [Model features](model_features.html) | Feature reference with summary and detail tables |
| Getting started | [EpiNow2](EpiNow2.html) | Quick introduction and basic usage |
| Infection model | [estimate_infections()](estimate_infections.html) | Mathematical model definition |
| Secondary model | [estimate_secondary()](estimate_secondary.html) | Mathematical model definition |
| Truncation model | [estimate_truncation()](estimate_truncation.html) | Mathematical model definition |
| Delay fitting | [estimate_dist()](estimate-dist.html) | Fitting delay distributions from linelist data |
| Gaussian process | [GP details](gaussian_process_implementation_details.html) | Kernel functions and spectral approximation |
| Practical workflow | [Workflow](estimate_infections_workflow.html) | End-to-end estimation and forecasting |
| Configuration examples | [Options](estimate_infections_options.html) | Different model configurations with results |
| Prior guidance | [Prior choice guide](prior_choice_guide.html) | Default priors and how to modify them |
| Production use | [epinow()](epinow.html) | Using `epinow()` and `regional_epinow()` |
| Case studies | [Case studies](case-studies.html) | External applications in the literature |
